Imperative Programming with Variational Effects

### 2018-09-10 paper/paper.tex
implement most of eric's feedback

\title{Imperative Programming with Variational Effects}
\author{Alex Grasley}
\degree{Master of Science}
\doctype{Thesis}
\department{Electrical Engineering and Computer Science}
\depttype{School}
\depthead{Director}
\major{Computer Science}
\advisor{Eric Walkingshaw}
\submitdate{August 27, 2018}
\commencementyear{2019}
\abstract{
Variation is a commonly encountered element of modern software systems.
Recent research into variation has led to increasing interest in the
development of variational programming languages and corresponding
variational models of execution. Variational imperative languages pose a
particular challenge due to the complex interactions between side effects and
variation. The development of interpreters for variational imperative languages
has produced a number of techniques to address these interactions. This
thesis builds upon and formalizes these techniques by defining a formal
operational semantics for a simple imperative language that supports both
variation and common side effects.

@@

While seemingly distinct, there is considerable overlap between these kinds of variation. Tooling or static analysis that deals with
statically resolved variation often encounters the same problems and employs the same techniques to solve them
that are found in variational execution. Chief among these issues is that brute-force techniques for dealing with variation quickly become infeasible as the
number of features grows. For a variational software system with $n$ features, the number of unique variants produced is $2^n$. Similarly, the naive approach
of running a program multiple times over varying inputs requires $2^n$ executions for inputs with $n$ features. This exponential growth poses what appears to
be an insurmountable obstacle to efficiently handling variation in the face of a large number of distinct features.

@@

The work in this thesis was primarily motivated by our work in defining a DSL for describing resource requirements in complex modern software
systems. The Resource DSL allows users to write programs that assert certain conditions of a successful resource configuration for a project.
Resource requirement specifications are an ideal setting for the use of variational execution. For many modern software systems there are a
whole host of configuration options available. Using a plain, non-variational model of execution a user must individually check each configuration that they wish to
test. As discussed above, if a user wishes to check an entire configuration it quickly becomes infeasible due to the exponential explosion of the number of times they must
rerun the checker. By contrast, the Resource DSL due to its fully variational model of execution is capable of exploiting the common computations shared among many different
possible configurations, leading to efficiency gains. In addition to the potential efficiency gains, directly modeling variation in our language allows for easier reasoning about
large configuration spaces. For instance, users can easily preconfigure a variational program to only run for certain features that they are interested in. Users can also more easily
query results across options, making it potentially easier to identify patterns for particular sets of features.

@@

This thesis provides two primary contributions: (1) a formal semantics for variational imperative languages
and (2) a description of our design of the Resource DSL. These are of course intrinsically linked, where
the work on formalizing the semantics of variational imperative languages serves as the foundation for
a specific variational imperative language, namely the Resource DSL.

@@

Chapter 4 (\emph{IMP}) describes the imperative language IMP.
We use IMP because its simplicity allows for the definition of the key
interaction between variation and side effects without unnecessary linguistic
clutter. We set out the syntax of IMP which we have extended to include catchable
exceptions, as well as a big-step operational semantics for IMP. This allows us
to have a useful point of comparison for our later work on variationalizing IMP.

@@

We say that a variational value is \emph{configured} when all choices
have been eliminated, yielding a plain value without choices. We call these resulting plain values
\emph{variants}. We define a function \emph{conf} which takes a
\emph{configuration} and a variational value and produces a variant. Under the view of dimensions as representing distinct features, a configuration for a formula tree gives
a boolean assignment for each dimension of the variational value being configured.
A simple way to do this is to define a configuration as a list of the features that are present, with all
other features assumed to be absent. In other words, configurations are a list of variables that should
evaluate to true, with all other variables evaluating to false. Configuring a formula tree is then simply evaluating
each condition given the variable assignment. If the condition evaluates to true, then the left variant
is chosen, otherwise the right variant is chosen.

Using boolean formulas as our tags provides several advantages. For example, we can simplify trees like
$\tagtree{B}{1}{\tagtree{A}{1}{2}}$ to the more compact form $\tagtree{(B \vee A)}{1}{2}$. This is just
one of a number of optimizations that are made possible by representing tags as boolean formulas
\cite{WO14gpce,HW16fosd}. Formula choices are also more convenient
for maintaining a variational context during execution of a program, a fact that will become
more relevant when we present our work on variational imperative programming later in
Chapter \ref{ch:vimp}. Despite these advantages, formulas incur some cost by introducing boolean
satisfiability problems into many common operations, which is an NP-complete problem. However, thanks to improvements in the
efficiency of modern SAT solvers, often the worst-case NP-complete performance of these satisfiability
problems can be avoided.

@@

% sharing is desirable. what is sharing?
The choice calculus enables several desirable properties for variational programs.
The first property that is possible to implement through the choice calculus \emph{sharing}. A naive approach to executing a variational program is simply to
configure the variational program in order to produce each variant and run them all sequentially.
For a program with $n$ dimensions this results in $2^n$ variants that must be executed. Crucially,
the naive approach must recompute any common elements shared between variants. Embedding
choices in our program, combined with a variability-aware model of execution allows us to exploit this
inherent sharing by only computing shared components once, greatly improving the efficiency of
executing a variational program.

@@

In \prog{vlookup} we throw a runtime exception when a variable is undefined. Our semantics does not define the behavior of trying to access an undefined variable, but
such cases are easily detected and prevented with a proper type system. Variational type systems have been studied extensively and readers seeking static guarantees
that undefined variable access is impossible are referred to those works \cite{kenner2010typechef,CEW12icfp,CEW14toplas}.

@@

This flag value has an interesting interaction with choices. We only wish for a choice statement to set its flag when all variants can safely be ignored. If there is a single variant that is still active, then a choice statement should not set its flag, in this way suppressing the flags of its children. When the flag is set from inside of
a choice, there is a simple procedure to check if the choice should propagate or suppress the flag. If both alternatives are either unsatisfiable or have the flag set, then
the parent choice should set its flag, indicating that execution is halted for all possible variants. In the Resource DSL, we implement this system via an error monad,
with errors serving as the flags and inserting catches for choices in order to inspect whether the parent should propagate the errors or suppress them.

@@

whenFalse = Not . whenTrue
\end{lstlisting}
\label{fig:when}
\caption{Control structure helper functions}
\end{figure}

@@

\subsection{Throwing Exceptions}

>>>>

\title{Imperative Programming with Variational Effects}
\author{Alex Grasley}
\degree{Master of Science}
\doctype{Thesis}
\department{Electrical Engineering and Computer Science}
\depttype{School}
\depthead{Director}
\major{Computer Science}
\advisor{Eric T. Walkingshaw}
\submitdate{August 27, 2018}
\commencementyear{2019}
\abstract{
Variation is a commonly encountered element of modern software systems.
Recent research into variation has led to increasing interest in the
development of variational programming languages and corresponding
variational models of execution. Variational imperative languages pose a
particular challenge due to the complex interactions between side effects and
variation. The development of interpreters for variational imperative languages
has produced a number of techniques to address these interactions. This
thesis builds upon and formalizes these techniques by defining a formal
operational semantics for a simple imperative language that supports both
variation and common side effects.

@@

There is considerable overlap between these kinds of variation. Tooling or static analysis that deals with
statically resolved variation often encounters the same problems and employs the same techniques to solve them
that are found in variational execution. Chief among these issues is that brute-force techniques for dealing with variation quickly become infeasible as the
number of features grows. For a variational software system with $n$ features, the number of unique variants produced is $2^n$. Similarly, the naive approach
of running a program multiple times over varying inputs requires $2^n$ executions for inputs with $n$ features. This exponential growth poses what appears to
be an insurmountable obstacle to efficiently handling variation in the face of a large number of distinct features.

@@

The work in this thesis was primarily motivated by our work in defining a DSL for describing resource requirements in complex modern software
systems. The Resource DSL allows users to write programs that assert certain conditions of a successful resource configuration for a project.
Resource requirement specifications are an ideal setting for the use of variational execution. For many modern software systems there are a
whole host of configuration options available. Using a plain, non-variational model of execution a user must individually check each configuration that they wish to
test. As discussed above, if a user wishes to check all configurations it quickly becomes infeasible due to the exponential explosion of the number of times they must
rerun the checker. By contrast, the Resource DSL due to its fully variational model of execution is capable of exploiting the common computations shared among many different
possible configurations, leading to efficiency gains. In addition to the potential efficiency gains, directly modeling variation in our language allows for easier reasoning about
large configuration spaces. For instance, users can easily preconfigure a variational program to only run for certain features that they are interested in. Users can also more easily
query results across options, making it potentially easier to identify patterns for particular sets of features.

@@

This thesis provides two primary contributions: (1) a formal semantics for variational imperative languages
and (2) a description of our design of the Resource DSL. These are of course intrinsically linked, where
the work on formalizing the semantics of variational imperative languages serves as the foundation for
our work on the Resource DSL.

@@

Chapter 4 (\emph{IMP}) describes the imperative language IMP \cite{winskel1993formal,nipkow1998winskel,nipkow2014concrete}.
We use IMP because its simplicity allows for the definition of the key
interaction between variation and side effects without unnecessary linguistic
clutter. We set out the syntax of IMP which we have extended to include catchable
exceptions, as well as a big-step operational semantics for IMP. This allows us
to have a useful point of comparison for our later work on variationalizing IMP.

@@

We say that a variational value is \emph{configured} when all choices
have been eliminated, yielding a plain value without choices. We call these resulting plain values
\emph{variants}. We define a function \emph{conf} which takes a
\emph{configuration} and a variational value and produces a variant. Under the view of dimensions as representing distinct features, a configuration for a formula tree gives
a boolean assignment for each dimension of the variational value being configured.
A simple way to do this is to define a configuration as a set of the features that are present, with all
other features assumed to be absent. In other words, configurations are a set of variables that should
evaluate to true, with all other variables evaluating to false. Configuring a formula tree is then simply evaluating
each condition given the variable assignment. If the condition evaluates to true, then the left variant
is chosen, otherwise the right variant is chosen.

Using boolean formulas as our tags provides several advantages. For example, we can simplify trees like
$\tagtree{B}{1}{\tagtree{A}{1}{2}}$ to the more compact form $\tagtree{(B \vee A)}{1}{2}$. This is just
one of a number of optimizations that are made possible by representing tags as boolean formulas
\cite{WO14gpce,HW16fosd}. Formula choices are also more convenient
for maintaining a variational context during execution of a program, a fact that will become
more relevant when we present our work on variational imperative programming later in
Chapter \ref{ch:vimp}. Despite these advantages, formulas incur some cost by introducing boolean
satisfiability problems into many common operations, which is an NP-complete problem. However, thanks to improvements in the
efficiency of modern SAT solvers, often the worst-case NP-complete performance of these satisfiability
problems can be avoided. Several studies have shown that the types of boolean formulas typically encountered in
variational software are efficiently solved by modern SAT solvers \cite{liebig2013scalable,Mendonca:2009:SAF:1753235.1753267}.

@@

% sharing is desirable. what is sharing?
The choice calculus enables several desirable properties for variational programs.
The first property that is possible to implement through the choice calculus \emph{sharing}. A naive approach to executing a variational program is simply to
configure the variational program in order to produce each variant and run them all sequentially.
For a program with $n$ dimensions this results in $2^n$ variants that must be executed in the worst case. Crucially,
the naive approach must recompute any common elements shared between variants. Embedding
choices in our program, combined with a variability-aware model of execution allows us to exploit this
inherent sharing by only computing shared components once, greatly improving the efficiency of
executing a variational program.

@@

In \prog{vlookup} we throw a runtime exception when a variable is undefined. Our semantics does not define the behavior of trying to access an undefined variable, but
such cases are easily detected and prevented by a type system. Variational type systems have been studied extensively and readers seeking static guarantees
that undefined variable access is impossible are referred to those works \cite{kenner2010typechef,CEW12icfp,CEW14toplas}.

@@

This flag value has an interesting interaction with choices. We only wish for a choice statement to set its flag when all variants can safely be ignored. If there is a single variant that is still active, then a choice statement should not set its flag, in this way suppressing the flags of its children. When the flag is set from inside of
a choice, there is a simple procedure to check if the choice should propagate or suppress the flag. If both alternatives are either unsatisfiable or have the flag set, then
the parent choice should set its flag, indicating that execution is halted for all possible variants. In the Resource DSL (Section \ref{sec:throwing}), we implement this system via an error monad,
with errors serving as the flags and inserting catches for choices in order to inspect whether the parent should propagate the errors or suppress them.

@@

whenFalse = Not . whenTrue
\end{lstlisting}
\caption{Control structure helper functions}
\label{fig:when}
\end{figure}

@@

\subsection{Throwing Exceptions}
\label{sec:throwing}
